---
title: "Durably Reducing Transphobia: Analysis of an Experiment"
subtitle: "POLS 4716"
date: "September 25, 2025"
output:
  html_document:
    theme: flatly
    highlight: tango
---

<!--
When you are ready, click on Knit button in R Studio to produce 
an HTML document of your results.
-->


Reference:
> Broockman, David and Joshua Kalla. 2016. "[Durably Reducing Transphobia: A Field Experiment on Door-to-door Canvassing.](https://science.sciencemag.org/content/352/6282/220.abstract)" *Science* Vol 352, No. 6282. pp. 220--224

The data being analysed in this noteboook comes from a randomized experiment conducted in December 2014 in South Florida on whether door-to-door canvassing with a short conversation can lastingly reduce intergroup prejudices. This particular experiment targeted antitransgender prejudice and 56 canvassers were sent door to door to encourage active perspective-taking with 501 voters. Each household was randomly assigned to either the active treatment of a conversation about transgender people or they were assigned a placebo conversation about recycling. Some canvassers identified as transgender while others identified as nontransgender.

## Codebook

We have taken a subset of their data and cleaned it up for use in today's session as `transphobia.csv`. Here are a list of the variables in the data set:

| Variable Name    | Description                                                                |
|------------------|----------------------------------------------------------------------------|
| `age`            | Age of the respondent in years                                             |
| `female`         | 1=respondent marked "Female" on voter registration, 0 otherwise            |
| `voted_gen_14`   | 1 if  respondent voted in the 2014 general election                        |
| `vote_gen_12`    | 1 if respondent voted in the 2012 general election                         |
| `treat_ind`      | 1 if respondent was assigned to treatment, 0 for control                   |
| `racename`       | character name of racial identity indicated on voter file                  |
| `democrat`       | 1 if respondent is a registered Democrat                                   |
| `therm_trans_t0` | 0-100 feeling thermometer about transgender people at baseline             |
| `therm_trans_t1` | 0-100 feeling thermometer about transgender people 3 days after treatment  |
| `therm_trans_t2` | 0-100 feeling thermometer about transgender people 3 weeks after treatment |


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

transphobia <- read_csv(url("https://tinyurl.com/trans-canvas-data"))
```

# 1. Descriptive Statistics

Let's start by looking at our data. Confirm the number of individuals in our dataset. Check how many individuals are in treatment vs. control. Look at the distribution of age in the data.

```{r look-at-your-data}
# TODO
```

## 1.1. Balance

Does it look like randomization worked? Check balance on pre-treatment variables between treatment and control units.

```{r calculate-balance}
# TODO
```

If you have extra time, display the balance in a barplot here.

```{r visualize-balance}
# TODO
```

# 2. Causal Effects

## 2.1. Estimating the SATE (after 3 days)

Using the method we discussed in class, produce our best estimate for the **sample average treatment effect (SATE)** of the treatment (perspective-taking) warmness towards transgender people 3 days after the experiment.

```{r sate-three-days}
# TODO
```

## 2.2. Estimating the SATE (after 3 weeks)

Now estimate the SATE on warmness towards transgender people 3 weeks after the conclusion of the experiment.

```{r sate-three-weeks}
# TODO
```


